---
title: Week 9 DraftKings Optimal Lineups
author: Aaron Miles
date: '2019-11-01'
slug: week-9-draftkings-optimal-lineups
categories:
  - Daily Fantasy
  - Fantasy
  - Football
  - Sports
  - Optimization
tags:
  - dfs
  - nfl
  - optimization
  - viz
keywords:
  - tech
---

We are BACK with optimal lineups from week 9 simulations!

Solid week last week, I broke about even, Kenny Golladay finally put on a show, and I think I'm about to manually exclude Tyler Boyd from being placed in lineups. 

No changes again, mainly due to work (that, you know, I get paid for), and working on Big Data Bowl, but I have been diving deep into modeling distributions, which I think can be a big help here.

(P.S. Sorry for all the spelling errors last week, yikes)

If you want to review the overall code for scraping and optimizing projections, the initial post is [here](https://amiles.netlify.com/2019/09/building-optimal-daily-fantasy-lineups-in-r/).

### Setup

```{r setup, error=FALSE, message=FALSE, warning=FALSE}
library(data.table)
library(dtplyr)
library(tidyverse)
library(rPref)
library(kableExtra)

week <- 9

proj <- readRDS(paste0('week_', week, '_proj.RDS'))

sal <- read_csv(paste0('DKSalaries_wk_', week, '.csv'))



```
I'll start with the optimized lineups pulled for week 4, with the same details as last time: 10,000 lineups, using the standard deviation of projections, completely individually based (still working on that).

```{r readLineups, error=FALSE, message=FALSE, warning=FALSE}

sim_lu <- readRDS(paste0('sim_lineups_week_', week, '.RDS')) %>%
  rename(pts_base=points) %>%
  select(lineup, Name, team, position, pts_base, pts_pred, sd_pts, Salary)

glimpse(sim_lu)


sim_lu %>%
  filter(lineup<=3) %>%
  arrange(lineup, position, desc(pts_pred)) %>%
  mutate_at(vars(pts_base, pts_pred, sd_pts), function(x) round(x, 2)) %>%
  knitr::kable() %>%
  kable_styling() %>%
  column_spec(1, bold=TRUE) %>%
  collapse_rows(columns = 1, valign = 'top') %>%
  scroll_box(height = '600px', width = '100%')
  

```

### Who is in Optimal Lineups?  

```{r top_presence, message=FALSE, error=FALSE, warning=FALSE, out.width='100%'}
sim_lu %>%
  group_by(Name, position, Salary) %>%
  dplyr::summarize(lu=n_distinct(lineup)) %>%
  ungroup() %>%
  group_by(position) %>%
  top_n(10, lu) %>%
  ungroup() %>%
  arrange(position, desc(lu)) %>%
  mutate(Name=factor(Name),
         Name=fct_reorder(Name, lu)) %>%
  ggplot(aes(x=Name, y=lu/1000, fill=Salary)) +
  geom_bar(stat='identity') +
  facet_wrap(~position, ncol = 3, scales='free') +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_viridis_c() +
  xlab('') +
  ylab('Lineups (Thousands)') +
  ggtitle('Top 10 Players Present by Position') 
```




Some of my observations: 

  
* Interesting week this week, given the uncertainty around the Chiefs QB Situation. Also, most weeks the model hammers DeShaun Watson and Lamar Jackson, and neither are on the main slate. This week we're hammering a lot of Russell Wilson. The 'buy low' options at QB are Philip Rivers and Ryan Fitzpatrick


* Lots of interchangability at the top of the RB pool, with McCaffrey, Chubb, Jones, Henry, and Bell all making string appearances.
  
* Keenan Allen showing up a lot at WR, hopefully he breaks out, as he's tweeted. Both Bucs WRs showing up as well. Model doesn't have us really paying up for WR this week
  
* It's the usual story at TE, there are only a few worth playing, but less segmented than usual, with Doyle and Gesicki bridging the gap. 
  
* Pathers, Redskins, and Dolphins at D again. As with most weeks, the model has selected the Defenses with the lowest prices.
  
### Who is getting placed in Lineups?

DraftKings provides scoring for `r nrow(sal)` players this week, but only `r length(unique(sim_lu$Name))` make it into optimized lineups. Why is that? To determine, I'll plot projected points vs salary, colored by whether or not they make it into optimized lineups, and sized by their projection standard deviation

```{r pts_vs_sd, error=FALSE, message=FALSE, warning=FALSE}
plyr_lu <- sim_lu %>%
  group_by(Name, position) %>%
  dplyr::summarize(lu=n_distinct(lineup)) %>%
  ungroup() 

proj %>% 
  filter(avg_type=='weighted') %>%
  mutate(Name = ifelse(pos=="DST", last_name, paste(first_name, last_name))) %>%
  inner_join(sal, by=c("Name")) %>%
  select(Name, team, position, points, Salary, sd_pts) %>%
  left_join(plyr_lu, by='Name') %>%
  replace_na(list(lu=0)) %>%
  mutate(lu_bin=ifelse(lu==0, '0 Lineups', '>=1 Lineups'),
         lu_5=cut(lu,5, labels = FALSE)) %>%
  ggplot(aes(x=Salary, y=points, color=lu_bin, size=sd_pts)) +
  geom_point() +
  scale_color_manual(values = c('red', 'blue'), name="") +
  geom_smooth(inherit.aes = FALSE, aes(x=Salary, y=points), method = 'lm', se=FALSE) +
  ylab('Projected Points') +
  xlab('Salary') +
  ggtitle('Who makes it into Optimized Lineups?') +
  scale_x_continuous(labels=scales::dollar)
```

This week, not many players below the line get into lineups, but some above get excluded due to their small uncertainty. Remember, this method takes players who have the potential to blow up rather than players with solid floors.


### Flex Configurations

In DFS lineups, you have an extra spot to use on an RB, WR, and TE of your chosing

```{r flex, message=FALSE, error=FALSE, warning=FALSE}
sim_lu %>%
  group_by(lineup) %>%
  mutate(lineup_pts=sum(pts_pred)) %>%
  group_by(lineup, position) %>%
  mutate(n=n()) %>%
  select(lineup, position, n, lineup_pts) %>%
  distinct() %>%
  spread(key=position, value=n) %>%
  filter(RB>=2, TE>=1, WR>=3) %>%
  mutate(flex=case_when(RB==3 ~ 'RB',
                        TE==2 ~ 'TE',
                        WR==4 ~ 'WR')) %>%
  group_by(flex) %>%
  dplyr::summarize(pts=median(lineup_pts),
                   cases=n()) %>%
  knitr::kable() %>%
  kable_styling(full_width = FALSE)

```

RB's lead flex configurations, WR's second, and TE's third.

### Pareto Lineups

```{r pareto, message=FALSE, error=FALSE, warning=FALSE}
lu_df <- sim_lu %>%
  group_by(lineup) %>%
  dplyr::summarize(lineup_pts=sum(pts_pred),
                   lineup_sd=sum(sd_pts)) %>%
  ungroup()

pto <- psel(lu_df, low(lineup_sd) * high(lineup_pts))


ggplot(lu_df, aes(y=lineup_pts, x=lineup_sd)) +
  geom_point() +
  geom_point(data=pto, size=5) +
  ylab('Lineup Points') +
  xlab('Lineup Points St Dev') +
  ggtitle('Lineup Points vs Uncertainty',
          subtitle = 'Pareto Lineups Bolded')
```


Here's a look at the pareto lineups.

```{r pareto_table, error=FALSE, message=FALSE, error=FALSE}
psel(lu_df, low(lineup_sd) * high(lineup_pts)) %>%
  left_join(sim_lu, by='lineup') %>%
  group_by(lineup) %>%
  arrange(lineup_pts, position, desc(Salary)) %>%
  select(lineup, lineup_pts, lineup_sd, Name, team, position, pts_pred, sd_pts, Salary) %>%
  mutate_at(vars(lineup_pts, lineup_sd, pts_pred, sd_pts), function(x) round(x, 2)) %>%
  knitr::kable() %>%
  kable_styling(fixed_thead = T) %>%
  column_spec(1:3, bold=TRUE) %>%
  collapse_rows(columns = 1:3, valign = 'top') %>%
  scroll_box(height = '500px', width = '100%')

```


Week 9 optimal lineups can be found [here](https://drive.google.com/open?id=1ANNV7ejZlYHYj8_T5Q8ul8rDcqdYj66T)